pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-02-28 article.tex
First draft ready for feedback by @rsetienne

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a), 
    that is converted to an alignment (2a), which is used
    to create a BEAST2 posterior of phylogenies (3a),
    which is compared to the true phylogeny (1a) \giovanni{this makes me think that we might need an arrow going from 1a to 4a, maybe (and i say maybe) the arrow could be in red to highlight that the "true" nltt is in red too. BTW, the red line is not supervisible.} using
    the nLTT statistic. (4a) shows the nLTT of the true tree in red,
    the nLTT of all individual posterior trees in grey, and the average
    nLTT statistic of all posterior trees in black.
    Taking the differences between the true tree's nLTT and each posterior's
    nLTT results in an error distribution, here displayed as a histogram
    of error values (5a).
    Optionally, a twin phylogeny (1b) can be generated from a true
    phylogeny (1a), which then results in an error distribution (5b) with
    the similar intermediate steps 
  }
  \label{fig:pipeline}
\end{figure}

@@

The second step selects an inference model, using the alignment \giovanni{Is it worth to add it in the figure, maybe on the side of the vertical pipeline? It could be something like an arrow moving horizontally to the box of inference model. Then the box could have another arrow pointing at the arrow connecting 2 and 3. If you think this is too complicated, and not easy to follow for the reader you can discard it.}.
The user can specify whether to use the generative model or select the best model from a set of inference models. 
When selecting the generative model, the site and clock model used in the alignment simulation are used in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

The third step infers a Bayesian posterior from the simulated 
alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a),
using the inference model(s) selected in the previous step. The user
can specify the additional parameters needed for the BEAST2 run, which
are the Markov-Chain Monte Carlo (MCMC) setup, 
an optional MRCA prior and an RNG seed \giovanni{This has been already said when defining the inference model. I suggest to define it only once in detail (like here) and just mention it the first time (something like: "An inference model is a combination of site model, clock model, tree prior as well as additional BEAST2 settings." in the description section, around line 98), which will probably look a bit more neat, as it's a bit clearer the parallel between alignment model and inference model.}.
The MCMC setup determines the number of posterior trees sampled.
An MRCA prior allows the inferred phylogeny to have a dated crown age.

The fourth step compares the true tree to each posterior tree
error using the nLTT statistic (\cite{janzen2015approximate}) 
(Fig.~\ref{fig:pipeline}, 3a $\rightarrow$ 4a), resulting
in an error distribution (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
\richel{I would not mind to remove 4a and 4b from the figure}
The nLTT statistic is used by default, but also a user-defined error statistic can be used. As an example,
\verb;pirouette; supplies one custom error statistic,
that uses an absolute difference in the gamma statistic [\cite{pybus2000testing}].
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of the state space.

@@

An optional step is to generate a 'twin tree' (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, 
yet its branching times are simulated according to 
the tree prior used by BEAST2 to produce a posterior \giovanni{is it true? is it Yule or BD? Please check}.
The goal of this procedure is to generate a tree 
to use as a control for the main experiment.
We produce the twin tree according to the following procedure.

@@

\begin{lstlisting}[
    language=R, floatplacement=H, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {Create a Yule tree}
  ]
phylogeny <- create_ideal_tree(n_taxa = 5, crown_age = 10)
\end{lstlisting}

\giovanni{do we really want to use a function called "create\_ideal\_tree"?}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_1.png}
  \caption{
    Example 1: the error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), the y-axis shows the error. \giovanni{This holds true for all the figures: if we want to add these numbers we need to specify what they represent in the caption.}
  }
  \label{fig:example_1}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_2.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), the y-axis shows the error.
  }
  \label{fig:example_2}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_3.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, picking the best inference model versus the control.
    Here, the twin column shows the error BEAST2 makes starting from the original tree versus the twin. 
    The x-axis specifies the tree type (which can be 'true' or 'twin'), the y-axis shows the error.
  }
  \label{fig:example_3}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_4.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative model is known, compared to candidates with similar inference models, in relationship to the background error.
    The x-axis specifies the tree type, the y-axis shows the error.
  }
  \label{fig:example_4}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
\centering
\begin{tabular}{ | r | l | l | l | l | }
\hline
\multirow{2}{*}{\textbf{Example}} & \multicolumn{2}{c|}{\textbf{True tree}} 
                                  & \multicolumn{2}{c|}{\textbf{Twin tree}} \\
\cline{2-5}
                                  & \textbf{median} & \textbf{range} & \textbf{median} & \textbf{range} \\
\hline
1  & $0.02$  & $0.0 - 0.27$  &         &               \\
2  & $0.02$  & $0.0 - 0.08$  &         &               \\
3  & $0.018$ & $0.0 - 0.1$   & $0.019$ & $0.0 - 0.11$  \\
4C & $0.22$  & $0.0 - 0.088$ & $0.009$ & $0.0 - 0.056$ \\
4G & $0.020$ & $0.0 - 0.075$ & $0.013$ & $0.0 - 0.052$ \\
\hline
\end{tabular}
\caption{Results of the examples. 
  For the true and twin tree,
  the median and range of the errors 
  is shown. Experiment 4 is split in two rows
  by either the candidate ('C') or generative model ('G'). \giovanni{How do you define range? Are we using the standard error? \url{https://en.wikipedia.org/wiki/Standard_error}}
}
\label{tab:results}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a), 
    that is converted to an alignment (2a).
    The experiments are filtered (3a), possibly based on the
    alignment (2a), to determine which experiments are actually run.
    The experiment that are actually performed, use the 
    alignment (2a) to create a BEAST2 posterior of phylogenies (4a),
    which is compared to the true phylogeny (1a) 
    using the nLTT statistic. (5a) shows the nLTT of the true tree in red,
    the nLTT of all individual posterior trees in grey, and the average
    nLTT statistic of all posterior trees in black.
    Taking the differences between the true tree's nLTT and each posterior's
    nLTT results in an error distribution, here displayed as a histogram
    of error values (6a).
    Optionally, a twin phylogeny (1b) can be generated from a true
    phylogeny (1a), which then results in an error distribution (6b) with
    the similar intermediate steps 
  }
  \label{fig:pipeline}
\end{figure}

@@

The second step selects experiments to be actually run, 
based on the experiments and the alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a) 
The user can specify whether to use the generative model or select the best model from a set of inference models. 
When selecting the generative model, the site and clock model used in the alignment simulation are used in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

The third step infers a Bayesian posterior from the simulated 
alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
using the inference model(s) selected in the previous step. The user
can specify the additional parameters needed for the BEAST2 run, which
are the Markov-Chain Monte Carlo (MCMC) setup, 
an optional MRCA prior and an RNG seed \giovanni{This has been already said when defining the inference model. I suggest to define it only once in detail (like here) and just mention it the first time (something like: "An inference model is a combination of site model, clock model, tree prior as well as additional BEAST2 settings." in the description section, around line 98), which will probably look a bit more neat, as it's a bit clearer the parallel between alignment model and inference model.}.
The MCMC setup determines the number of posterior trees sampled.
An MRCA prior allows the inferred phylogeny to have a dated crown age.

The fourth step compares the true tree to each posterior tree
error using the nLTT statistic (\cite{janzen2015approximate}) 
(Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a), resulting
in an error distribution (Fig.~\ref{fig:pipeline}, 5a $\rightarrow$ 6a).
The nLTT statistic is used by default, but also a user-defined error statistic can be used. As an example,
\verb;pirouette; supplies one custom error statistic,
that uses an absolute difference in the gamma statistic [\cite{pybus2000testing}].
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of the state space.

@@

An optional step is to generate a 'twin tree' (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, 
yet its branching times are simulated according to 
the tree prior used by BEAST2 to produce a posterior 
\giovanni{is it true? is it Yule or BD? Please check}
\richel{It is Yule}.
The goal of this procedure is to generate a tree 
to use as a control for the main experiment.
We produce the twin tree according to the following procedure.

@@

\begin{lstlisting}[
    language=R, floatplacement=H, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {Create a Yule tree}
  ]
phylogeny <- create_yule_tree(n_taxa = 5, crown_age = 10)
\end{lstlisting}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_1.png}
  \caption{
    Example 1: the error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), 
    the y-axis shows the error. 
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_1}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_2.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
    The x-axis specifies the tree type (which can be 'true' or 'twin'), the y-axis shows the error.
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_2}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_3.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, picking the best inference model versus the control.
    Here, the twin column shows the error BEAST2 makes starting from the original tree versus the twin. 
    The x-axis specifies the tree type (which can be 'true' or 'twin'), the y-axis shows the error.
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_3}
\end{figure}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_example_4.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative model is known, compared to candidates with similar inference models, in relationship to the background error.
    The x-axis specifies the tree type, the y-axis shows the error.
    The boxplot shows the minimum, first quartile, median, third 
    quartile and maximum error, of which the values are displayed 
    in the adjacent label
  }
  \label{fig:example_4}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
\centering
\begin{tabular}{ | r | l | l | l | l | }
\hline
\multirow{2}{*}{\textbf{Example}} & \multicolumn{2}{c|}{\textbf{True tree}} 
                                  & \multicolumn{2}{c|}{\textbf{Twin tree}} \\
\cline{2-5}
                                  & \textbf{median} & \textbf{range} & \textbf{median} & \textbf{range} \\
\hline
1  & $0.02$  & $0.0 - 0.27$  &         &               \\
2  & $0.02$  & $0.0 - 0.08$  &         &               \\
3  & $0.018$ & $0.0 - 0.1$   & $0.019$ & $0.0 - 0.11$  \\
4C & $0.22$  & $0.0 - 0.088$ & $0.009$ & $0.0 - 0.056$ \\
4G & $0.020$ & $0.0 - 0.075$ & $0.013$ & $0.0 - 0.052$ \\
\hline
\end{tabular}
\caption{Results of the examples. 
  For the true and twin tree,
  the median and range (from minimum to maximum value) of the errors 
  is shown. Experiment 4 is split in two rows
  by either the candidate ('C') or generative model ('G')
}
\label{tab:results}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
